---
title: "looping and file selection"
author: "P,H, Whitfield"
date: "`r format(Sys.time(), '%Y-%B-%d')`"
output:
  bookdown::html_document2:
    number_sections: yes
    theme: "cerulean"
    toc: yes
    toc_float: yes
  bookdown::pdf_document2:
    toc: yes
    number_sections: yes
  bookdown::word_document2:
    toc: yes
---

# Load required libraries

Run the code chunk below to load the libraries required for this workbook.

```{R, warning = FALSE, message = FALSE}
library(knitr)
library(bookdown)
library(here)

library(CSHShydRology)
library(tidyhydat)
library(whitebox)

# wtb_init(exe_path='C:/Users/paulh/Documents/R/win-library/4.1/whitebox/WBT/whitebox_tools.exe')

hy_set_default_db("C:/Program Files (x86)/ECCC/ECDataExplorer/Database")
```

# Introduction

This document demonstrates some of the CSHShydRology package functions that deal with examining the quality of the data.

# getting file lists  


```{r}

rd <- here::here()
rd

# get the names of files containing flow data
files  <- list.files(here::here("data"))

files


```


So, we see this directory contains files of different types and we only want the files from ECDE.



```{r}


# get the names of files containing flow data
files  <- list.files(here::here("data"), pattern = "Daily_Flow_ts.csv")

files


```
```{r, echo=TRUE}

f_in <- read.csv(here::here("data",files[1]))

str(f_in)

head(f_in)

tail(f_in)


```

We see several things. 

1. Date is character
2. There are three lines without data, but with a Notice about the data.  We get around this by using `ch_read_WSC_flows()`

```{r, echo=TRUE}

f_in <- ch_read_ECDE_flows(here::here("data",files[1]))

str(f_in)

head(f_in)

tail(f_in)


```

# looping over files

Now we cna use a loop to produce a plot for each file and get summary information into a data frame.  A key thing it to think about having a link in the new dataframe that can be used to connect to other dataframes such as those that contain metadata or other results.  Here we used the stationID.

```{r, echo=TRUE}
nn <- 13
result <- array(NA, dim = c(length(files), nn))

for (kk in 1:length(files)) {
     
     f_in <- ch_read_ECDE_flows(here::here("data",files[kk]))
   
     result[kk,1]    <-  f_in[1,1]  #stationID
     result[kk,2:7]  <-  summary(f_in$Date)
     result[kk,8:13] <-  summary(f_in$Flow)
     
  # generate a raster plot of the flow data
     ch_flow_raster(f_in)
}

result <- data.frame(result)

colnames(result) <- c("StationID", "Date_min", "d_25", "d_median", "d_mean", "d_75", "Date_max",
                     "Q_min", "Q_25", "Q_median", "Q_mean", "Q_75", "Q_max")

print(result)

write.csv(result,here::here("output","sample.csv"))

```

# other functions

It is easy to see what other functions are available in CSHShydRology. The following will go to the main help page.


At the bottom of the help screen you will see a link to "Index". Follow that and you will get a list of all the package help pages.  


Now modify the looping structure above to produce regime plots for the first five files.


