---
title: "0_dataframes"
author: "Paul Whitfield"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(CSHShydRology)
```

## dataframes

Dataframes are a common two-dimensional data structure in `R`.  Technically, a data.frame is a list of vectors of equal length, and we normally think of this as some form of table with columns and rows.  Cases are the rows and variables the columns.  



```{r, echo=TRUE}

name <- c("Dan", "Kevin", "Robert", "Jason", "Paul")
age <- c(61, 61, 32, 25, 70)
home <- c("Galiano", "Saskatoon", "Waterloo", "Coquitlam", "Canmore")

df <- data.frame(name, age, home)

class(name)
class(age)
class(df)

print (df)

```
# alternative method

Here is an alternative way to create the data.frame.  In R there are usually multiple ways to get to the same place, but some are slower / faster.

```{r, echo=TRUE}

df1 <- data.frame(
  name = c("Dan", "Kevin", "Robert", "Jason", "Paul"),
  age = c(61, 61, 32, 25, 70),
  home = c("Galiano", "Saskatoon", "Waterloo", "Coquitlam", "Canmore")
)

print(df1)

```




# find out about the data.frame

Rather than listing the entire data.frame we can get details that could help understand the data.

```{r, echo=TRUE}

names(df)

ncol(df)
nrow(df)

typeof(df)
length(df)

class(df)
# so a data.frame is a special case of a list

str(df)


```


# factors

Need to know about factors.  Factors are useful grouping variables.  There are caes where strings including dates are interpreted as factors and that can create problems.


```{r, echo=TRUE}

test <- c("good", "good", "bad", "bad", "bad")

df <- data.frame(df,test)
df <- data.frame(df,test)

str(df)

df$test <-as.factor(df$test)

str(df)


```



# accessing components of a dataframe

We can use several operators to access columns [, [[, or $.  There is a differnce in the result: [ gives a dataframe, the others a vector.

```{r, echo=TRUE}

df["name"]

df$name

df[["name"]]

df[[1]]


```
# access as a matrix

reference to row then column

Notice that the reference difference between [2,3], and [2,3,]



```{r, echo=TRUE}

df[2,3,]

df[df$age >= 50,]

df[2:5,3]

df[2:5,3,]

df[2:5,3, drop = FALSE]

df[-(2:3),]

## use of NA

df[2,3] <- NA

df

df[!is.na(df$home),]


```

# adding to dataframe


add a row with rbind [bind rows]

add a column with cbind [column bind] or by naming a new column df$xxx

add a column using data.frame

Note: these changes need to be "assigned" using "<-".

```{r, echo=TRUE}

df1 <- df1[1:5,1:3]
rbind(df1,list("Tom", 44, "Toronto"))
df1

df2 <- rbind(df1,list("Tom", 44, "Toronto"))
df2

df1 <- cbind(df2,prov = c("BC","SA","ON","BC", "AB","ON"))

prov2 <-c("BC","SA","ON","BC", "AB","ON")

new_df <- data.frame(df2,prov2)

```

# deleting 

removing rows and columns and changing data points

```{r, echo=TRUE}

str(df1)

df1$prov <- NULL


df1

# similarly a row

df1 <- df1[-6,]

df1

df1[df1  == 61] <- NA

df1

```


# an aside on `str`

`str` can be used on just about anything.

```{r, echo=TRUE}
data("CAN05AA008")

str(df)
str(CAN05AA008)
str(mean)
str(ch_binned_MannWhitney)


```



# an aside on "::" and ":::"


You can reference an external package function directly without loading the package using "::"
You can reference an internal package function directly without loading the package using ":::"


```{r, echo=TRUE}

Kendall::MannKendall

# Kendall::summary.Kendall

Kendall:::summary.Kendall
```
