---
title: "Introduction challenge exercises"
author: "Kevin Shook"
date: "June 5, 2022"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Challenge 1

Using the `ifelse` function, create a new column "hydroyear"
in the CAN01AD002 data frame. If the month is 10 or greater,
the hydro year is one greater than the year. Otherwise, it is the
current year. Hint: you can use the `format` command to get the month
from the date.

```{r}
library(CSHShydRology)
CAN01AD002 <- CAN01AD002
CAN01AD002$year <- as.numeric(format(CAN01AD002$date, format = "%Y"))
CAN01AD002$month <- as.numeric(format(CAN01AD002$date, format = "%m"))
CAN01AD002$hydroyear <- ifelse(CAN01AD002$month < 10, CAN01AD002$year, CAN01AD002$year + 1)
summary(CAN01AD002)
```

## Challenge 2

Get the mean daily discharge for each hydro year. Plot the values.
```{r echo = TRUE}
hydro_year_mean <- aggregate(flow~hydroyear, data = CAN01AD002, FUN = "mean")
summary(hydro_year_mean)
```
### Using standard plotting
```{r echo = TRUE}
plot(hydro_year_mean, xlab = "Hydro year", ylab = "Mean daily discharge (m続/s)",
     main = "WSC Gauge 01AD002", type = "l")
```
### Using ggplot2
```{r echo = TRUE}
library(ggplot2)
p <- ggplot(hydro_year_mean, aes(hydroyear, flow)) +
  geom_line() +
  xlab("Hydro year") +
  ylab("Mean daily discharge (m続/s)") +
  ggtitle("WSC Gauge 01AD002")

p

```

## Challenge 3

Classify the daily flows by quarter: Q1 = Jan-Mar, Q2 = April-June,
Q3 = July-September, Q4 = October-December.
Plot histograms of the daily flows facetted by each quarter using **ggplot2**
with the function `facet_wrap`.

### Solve using `ifelse`

```{r echo = TRUE, warning=FALSE}

CAN01AD002$quarter <- ifelse(CAN01AD002$month <= 3, "Q1", 
                             ifelse(CAN01AD002$month <= 6, "Q2", 
                                    ifelse(CAN01AD002$month <= 9, "Q3", "Q4")))
summary(CAN01AD002)
```

### Plot using **ggplot2**

```{r echo = TRUE, warning=FALSE, message=FALSE}

p <- ggplot(CAN01AD002, aes(flow)) +
  geom_histogram() +
  xlab("Daily discharge (m続/s)") +
  ylab("Number of days") +
  ggtitle("WSC Gauge 01AD002") +
  facet_wrap(~quarter) 
p
```

### Solve using factors

```{r}
CAN01AD002$quarter <- ifelse(CAN01AD002$month <= 3, "Jan-Mar", 
                             ifelse(CAN01AD002$month <= 6, "Apr-Jun", 
                                    ifelse(CAN01AD002$month <= 9, "Jul-Sept", "Oct-Dec")))

CAN01AD002$quarter <- factor(CAN01AD002$quarter, 
                             levels = c("Jan-Mar", "Apr-Jun", "Jul-Sept", "Oct-Dec"))
summary(CAN01AD002)
```
### Plot using **ggplot2**

```{r echo = TRUE, warning=FALSE, message=FALSE}

p2 <- ggplot(CAN01AD002, aes(flow)) +
  geom_histogram() +
  xlab("Daily discharge (m続/s)") +
  ylab("Number of days") +
  ggtitle("WSC Gauge 01AD002") +
  facet_wrap(~quarter) 
p2
```
