---
title: "Introduction to R Tutorial"
author: "Kevin Shook and R. Chlumsky"
date: "June 5, 2022"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to R Tutorial

This tutorial is intended to provide some background training on using **R** 

- Getting help in R
- Basic R Syntax
- Data types
- Flow control
- Creating plots (base R)
- Advanced topics
- Challenge exercises


## How this document works

This document was generated using **[R Markdown](https://rmarkdown.rstudio.com/articles_intro.html)**. This is a great tool for code transparency and data analysis, because the code blocks, code outputs, and your comments are "knit" into a single document! 

This document has a number of headings, text blocks (like this one), and code chunks that are run as the document is compiled. This allows you to see the script and output in one document (plus the code gets run and checked for errors as the document is made, which helps to reduce typos in the document code sections).

If you want to run one line in the chunk you can place your cursor in one line and press CTRL+ENTER to run that line (CMD+ENTER for Mac). Here's an example with just one line

```{r comment example}
# this is a comment in R. It is preceded by an octothorpe (or pound symbol, 
# or hashtag, depending on your generation).

print("This is a line of code with text in it, preceeded by ## in the output printout.")
```


Note: when you see the pound sign (#, aka octothorpe or "hash") next to a line of code, what follows to the right is a text comment that does not affect the code. Commenting is an important part of remembering what complex lines of code are meant to do!

## Where are you working?

One of the most common issues that new R users encounter is understanding the working directory. Your working directory is the folder location where your R session is operating from, and is where all files are assumed to be (and written to) unless otherwise specified. You can check your working directory with the command `getwd()`. You can also change the working directory with `setwd()`. Alternatively, if you are working in RStudio, your Files window can be used to view Files and navigate the working directory (same Pane where you can find Plots/Packages/Help/Viewer). If you click on *Files --> More*, you can see options to set or go to your working directory. Take a moment to check this before proceeding with this document.

## Getting Help in R

One of the first things to know is where to go for help. All functions that you come across in **R **  have documentation and examples to help you understand them. They can be accessed with either the `help()` function (e.g. `help(rep)`), or by using a question mark in front of the function name (e.g. `?rep`). 

```{r getting help, eval=TRUE}
# get help on command, like the variance function 'rep()'
help(rep)
?rep
```

If you don't quite know the name of a function, two "?" question marks followed by a guess will search for all the related topics. 

To search for all the help on calculating the variance, try this
```{r}
?? variance
```


Finally, if you place your cursor within the function name in your chunk and press **F1** on your keyboard, the documentation will again come up. 

If you are using **RStudio** , the help information should appear on the right side of your screen in the Help menu. If you don't see the Files/Plots/Help menu, then at the top menu, click on *View --> Panes --> Show All Panes*, and your layout should be restored.


For more resources, check out the CRAN website (<https://cran.r-project.org/>) for information, documentation, etc. 

It's hard to Google "R". To search for help on **R** topics use <https://rseek.org/>.



\newpage
## Basic R Syntax {#basicsyntax}

### Basic Arithmetic

Let's try some basic arithmetic. The `*` symbol is used for multiplication, 
`/` is used for division, and `^` is for exponentiation (raising to powers).

```{r basic arithmetic}
1 + 1
3 - 1
2 * 2
4 / 3
3 ^ 2
```

### Variables
Like all programming languages, **R** allows you to store values in named variables.
**R** is not strict in its definition of data types, so you can assign and re-assign the data types to variables at your whim.

The symbol (<-) *assigns* the value from one end of the arrow to the object at the point of the arrow. This method is directional, but you can also use the equal sign (=) as a left-only equivalent. Note that using <- is considered to be better style in **R** (the reasons are complicated), so you should be using it and that the
<- symbol is only used to assign values to variables, while the = sign is also used to specify parameter values to
functions. In *Rstudio*, you can insert <- by pressing [Alt][-].

Variable names have to begin with a letter, and can't include spaces or other sybols, other than
numbers, periods (.) and underscores (_). There are
also some reserved names that you can't use: 
`if else repeat while function for in next break TRUE FALSE NULL Inf NaN NA NA_integer_ NA_real_ NA_complex_ NA_character_` as they are used by **R** itself.

**R** is sensitive to case. So the variables `my_variable` and `My_variable` are completely different.
```{r variable assignment1}
my_variable = 1
My_variable <- 5
my_variable
My_variable
```
Coming up with good names for variables can be difficult. This document
discusses some of the naming conventions that are in use:
<https://journal.r-project.org/archive/2012-2/RJournal_2012-2_Baaaath.pdf>


### Functions

**R** uses functions, which are pre-written pieces of code, for everything.
Functions are called using round brackets (parentheses). Inside the brackets you
specify the values (often called arguments) used by the function.
In this example, the `rnorm` function is called, followed by the `mean` and `sd` (standard deviation) functions.
```{r running commands}
a <- rnorm(n = 20, mean = 0, sd = 2) # sample a normal distribution of 20 values with mean = 0, sd = 2
mean(a)                              # calculates the mean of values in vector a
sd(a)                                # calculates standard deviation of a
```

You can create your own **R** functions, which is very useful as it makes your
work even more reproducible. The functions can
be stored in an **R** package


### Checking data types

Here are some commands to check what type of data you are dealing with, and help you convert between different types.

```{r data types}
# what class of variable is it?
b <- 6
class(b)

# what 'typeof' value is b?
typeof(b) # 'double' is the type of number (double precision) 
```

```{r}
b <- "hello, world" # change the value assigned to b to a character value
b

typeof(b) # this has now changed from numeric/double
```

Convert between data types
```{r}
as.numeric("501") + 1  # convert quoted string to numeric and add 1
round(5.6)             # round to integer
floor(5.6)             # round down to integer
as.character(502)      # convert numeric to character
```

Now that we've started re-assigning values to variables, it's useful to recall their new values. On the upper-right portion of your screen in RStudio, the **Environment** tab shows the values of all user-defined variables and functions.

### Working with Text

R can read, write, and modify strings of text. A character string is any sequence
of any characters. Character strings are denoted by being enclosed in either
single `''` or double `""` quotes.

```{r}
a <- "This is a character string"
b <- 'This is another one'
a
b
```
You can't do arithmetic with character strings, but they are still very useful. Many types
of data, from names of places to names of months are characters. **R** has many functions
for working with character data. You can sort character strings, slicem them into pieces,
or combine them together.

You can combine strings together using the `paste` function, which concatenates them
together. Note that by default `paste` will add a space between character strings. If you
want to override this or use another separator, you have to specify it.

```{r}
day_num <- "05"
month_num <- "06"
year_num <- "2022"
date <- paste(day_num, month_num, year_num, sep = "/") 
date
```

### Dates
**R** is excellent at working with dates (and times). There is a built-in `date` data
type. You can convert a character date to a `date` type using the function `as.Date()`.
`date` values can be used for arithmetic.

If you don't tell `as.Date` the format, it will guess. However the guess may be wrong. You can specify the format using
the symbols shown below, which represent the day, month and year, as well as the "/" symbol which separates them. If you want to see all of the available formatting symbols, use the
command `? strptime`.
```{r}
day1 <- as.Date("1900-01-01")
day2 <- as.Date("05/06/2022", format = "%d/%m/%Y")
interval <- day2 - day1
interval
```


## Vectors, data frames, lists, arrays and matrices

A single valued variable, such as the ones above (also known as a scalar), is called an *atom* . Multiple atoms come together to form a set of values, called a **vector**. Multiple vectors form a **matrix** or **data frame**.

`c()` is the concatenate function, taking multiple atoms separated by commas and creating a new vector. 
We saw above how we can use addition or multiplication on an atom. We can also perform a single operation on all values in a vector. Let's perform apply arithmetic on the new vector `a`.
```{r vector operations1}
a <- c(1,2,3,4,5)
a
```


### Vectorized functions
Most functions in **R** are _vectorized_ meaning that they are applied
over a whole vector. For example the `paste` function works this way.
In this example, the paste function is applied to each value in the 
vector `dates`. Because there is only one value of `year` the value
is _recycled_ for each value in `dates`. This is really powerful.

```{r}
year <- "2020"
dates <- c("01/01", "02/01", "03/01", "04/01", "05/01")
complete_dates <- paste(dates, year, sep = "/")
complete_dates
```

The same applies to arithmetic
```{r vector operations2}
a <- seq(1, 10)
a
b <- sqrt(a)       # uses built-in square root function on all values of a
b
```

### Subsetting Vectors

It's important to be able to subset vectors, to extract specific values. To do this, we use the square brackets "[ ]". They are not to be confused with () or {}, which each have their own special uses, as discussed above. The `length` function returns the length of a vector.


```{r subsetting vectors 1}
# set up a vector
# sequence from 1 to 30, jumping by a value of 2 (instead of the default 1)
a <- seq(1, 30, by = 2)
a
length(a) # get the length of vector a (number of values)
```

```{r subsetting vectors 2}
# subset by location
a[1] # first value in the vector
a[c(1,2)] # first and second value in the vector
a[c(1,2,3)] # note the use of c() to specify multiple values
a[1:3] 
a[-1]  # all values without the first one
a[c(-1, -2, -3)] # omits the first 3 slements
```

### Using Booleans

Boolean operators refer to greater than (>), less than (<), or equal to (==).
Other operators are greater than *or* equal to (>=), and not equal to (!=)
```{r boolean checks for atoms}
1 > 3
1 < 3
1 <= 3 # less than or equal to
1 != 3
1 == 1
```
Note: **<=** looks very similar to **<-** but they are nothing alike.

We can also apply these basic operations with vectors, as shown below. In **R**, this will automatically apply the function `>` for all elements provided.

```{r boolean checks for vectors} 
c <- seq(1,10)
c > 5
```

You can use Booleans to perform subsetting operations (i.e. extract values by conditions).
```{r boolean subsetting1}
# set up vectors
a <- seq(1,10,2)
b <- seq(2,10,2)
a
b
```

```{r boolean subsetting2}
# subset vectors
a > 5 # which values are greater than 5?
a[ a > 5] # same as above, but returns the values of a that meet the criteria
a[(a >= 3) & (a < 8)] # multiple conditions for subsetting, take values >=3 and <8 only
subset(a, a >= 3 & a < 8) # same operation with the subset function
a[which(a >= 3 & a < 8)] # same operation as above with the which function

which( a >= 3 & a < 8) # returns the index values *which* meet the criteria specified
any(a %in% b)  # check if any values of a exist in b (FALSE, since we have evens and odds)
```


### Data frames

Data frames are the most useful data structure in **R** and are created by assembling 
vectors. Each vector becomes a column in the data frame, and can be any data type, which 
is what makes data frames so useful. The command `data.frame` is used to construct
a data frame from individual vectors. Note that all of the vectors must have the same
number of values. The function `length()` is useful to get the number of values
in a vector.

In this example we create 2 vectors, one containing dates and the other containing
numeric values, and put them in the same data frame. 

```{r create vectors and datrame}
date <- seq(day1, day2, by = 1)    # a sequence between the 2 dates defined earlier 
value <- runif(length(date))       # random numbers of the same length as 'date'

df <- data.frame(date, value)      # assemble the data frame
summary(df)                        # summarize the values in the data frame
```

If you click on `df` in the **Environment** tab, you can see the values.
This is _really_ useful.

#### Subsetting data frames
The same principles that work on selecting and subsetting vectors also work
with data frames (and arrays and matrices).
Note that the `length()` function is not useful as it just returns the number of
variables(columns) in a data frame.

```{r data frame and matrix}
length(df)
dim(df)    # returns [number of rows, number of columns]
nrow(df)   # number of rows
ncol(df)   # number of columns
```

Subsetting data frames, as with vectors, uses the square brackets "[ ]", but instead of one value, you list the desired row and column, separated by a comma. Note that the row is specified first, and the column second. Calling a row of a dataframe will return a mini dataframe with just the one row
```{r matrix reference}
df[2, 2]   # second row, second columns

df[2,] # leaving an entry blank will return the entire row or column specified
 
```

You can select any number of rows and/or columns
```{r more data frame references}
df[1:3, 2]                # rows 1-3, column 2
df[4:8,]                  # rows 4-8, all columns
df[c(1,3,5), 1]           # rows 1,3,5, column 1
```

You can also subset data frames by specifying the column names using "$"
The function `names()` is used to list the names in a data frame and to set them.
```{r data frame subsetting by variable}
names(df)
df$date[1:3]
df2 <- df
names(df2) <- c("DATE", "VALUE")
names(df2)
names(df2)[1] <- "Date"
names(df2)
```
#### Examples of working with data frames

The CSHShydRology package has some built in data that we can use when
we load the package
```{r}
library(CSHShydRology)
data()
```

**R** uses 'lazy-loading' so the data isn't loaded until you ask for it.
You can see that the data frame has 2 columns, the date and the flowrate.
```{r}
CAN01AD002 <- CAN01AD002
summary(CAN01AD002)
```

Because data frames hold more than one column of data, you can select values from
one column based on another.

For example, this would calculate the mean flows between 2 dates. The
data frame is subset by selecting rows where the `date` variable lies between
the start and end dates. The mean flow can then be calculated using the 
`mean` function on the specified column.

```{r}
start_date <- as.Date("2014-01-01", format = "%Y-%m-%d")
end_date <- as.Date("2014-12-31", format = "%Y-%m-%d")
selected <- CAN01AD002[CAN01AD002$date >= start_date & CAN01AD002$date <= end_date,]
summary(selected)
mean_flows <- mean(selected$flow)
cat("mean daily flow:", mean_flows, "m3/s")
```

What if you wanted the maximum or minimum yearly values?
One way would be to add a column for the year of each date. You can
do this using the `format()` function. Because `format()` returns a character
string, it's often more useful to convert it to a number, using the function
`as.numeric()`.

```{r}
CAN01AD002$year <- as.numeric(format(CAN01AD002$date, format = "%Y"))
summary(CAN01AD002)
```

Having added the years, you can get the summary statistic for each year using the
function `aggregate()`. The tilde (~) is used in **R** when you want
to express a relationship between two variables. In this case we are saying
that we want to aggregate the flows with respect to the years.

```{r}
yearly_max <- aggregate(flow~year, data = CAN01AD002, FUN = "max")
yearly_max
```

### Sorting
The best way to sort data (vectors and data frames) in **R** is using the `order` function. It sorts values, but returns their _order_, not the sorted values. It may seem strange, but is actually
very powerful.

In this example, we will sort the annual flows from largest to smallest. What you are seeing
is the _location_ (i.e. its row) of each value in its column, in order from largest to
smallest.

```{r}
order(yearly_max$flow, decreasing = TRUE)  # increasing order is the default
```

If you want to see the values arranged by largest to smallest, the command would be
```{r}
yearly_max$flow[order(yearly_max$flow, decreasing = TRUE)]
```

If you want to see the years arranged by largest to smallest max flows, the command is
```{r}
yearly_max$year[order(yearly_max$flow, decreasing = TRUE)]
```

### Factors

**R** has an unusual data type, **factor**, which is useful for dealing with
categorical data. Factors are only found is vectors, or data frame columns.
Although factors look like characters, they are actually
integers. Factors are useful for grouping other data values. In this
example, we have 4 hydrometric stations whose status us either "A" (active) or
"D" (discontinued). We use the `factor` function to assign factor levels
to each status. The `summary` command shows the number of values for each
level.

```{r}
station <- c("05HF021", "05HG001", "05HG002", "05HG003")
status <- c("D", "A", "A", "D" )
status <- factor(status, levels = c("D", "A"))
station_df <- data.frame(station, status)
summary(station_df)
```

You can select the values by treating them as characters
```{r}
station_df[station_df$status == "D",]
```

You can also sort them using the `order` function.  Because we defined the levels
of the factors to be "D" before "A", we can sort the values in this order. This is
very useful when you want to override the default alphabetical sorting order on
graphs.

```{r}
station_df[order(station_df$status),]
```


### Lists

Lists are one of the most powerful ways of storing data in **R** because they
can contain _any_ type of data - including other lists!

A list is created with the command `list()`. You can specify the names of
the list elements when it is created. In this example, the list contains
characters, a numeric value and a data frame.
```{r}
gauge_info <- list(number = "01AD002", name = "SAINT JOHN RIVER AT FORT KENT", prov = "NB",
                   drainage_area = 14700, annual_max_daily_Q = yearly_max)
names(gauge_info)
```

To add or extract values from a list, you can select the variable by its name (if it has one),
or by its number in the list, using `[[]]` (double square brackets).

```{r}
gauge_info[[1]]
gauge_info[[3]]
gauge_info$name
```

### Matrices and arrays

Because data frames are so useful and powerful, **R** uses arrays and matrices
less frequently. However, they can still be useful and some functions
will require or return arrays/matrices, so you need to know about them.

Things you need to know:  

1. arrays and matrices can only contain 1 type of data  
2. an array can have any number of dimensions, a matrix is only two-dimensional  
3. some functions like `t()` only work on matrices, not arrays  
4. you can convert 2d arrays and matrices to data frames and data frames with a single type
to arrays and matrices.  

The `matrix()` function creates a matrix. You can give all the matrix elements
an initial value. Note that the values are assigned one column at a time.

```{r}
# d <- matrix(base value for matrix, number of rows = 3, number of columns = 4)
a <- seq(1, 12)
d <- matrix(a, nrow = 3, ncol = 4)
d
typeof(d)
dim(d)
```

**R** has all of the usual matrix manipulation and math functions. One of the most
useful is `t()` which transposes the rows and columns.

```{r}
t(d)
```


## Flow control

It's often useful for **R** to control the order and number of statements which
are executed. You do this with the `if` statement and loops. These statements
can help you when you want to process a lot of data.

### Branches (if statements)
The `if` statement consists of a test statement, which returns
a Boolean value (`TRUE` or `FALSE`). If the result is `TRUE` then the
following statment is executed. If it is `FALSE` then nothing is done, unless
there is an `else` statement. In that case, the statement following the `else`
would be executed. If you want the `if` to execute more than one line after
the test, enclose the lines in braces `{}`.

```{r if statement}
a <- 5
if (a < 2) {
  print("a is less than 2!")
} else {
  print("a is not less than 2!")
}
```

Here the printed statement indicates that "a is not less than 2", since we defined `a` with a value of 5. This same code would print the other statement if we instead gave `a` a value less than 2.


### ifelse

Often you will want to apply an `if` to each element of a vector or a data frame column.
The simplest way to do this is with the `ifelse` command, which has the syntax
`ifelse(test, TRUE value, FALSE value)`. The test is applied to each element and either the TRUE value
or the FALSE value is returned for each.


You can do very useful things with `ifelse` on dataframes. For example, you could classify the flows
based on their value. In this example, the flows are tested to see if they are within
the lowest decile and a comment "low flow" is added if they are.
```{r}
decile <- quantile(CAN01AD002$flow, 0.1)
decile
CAN01AD002$comment <- ifelse(CAN01AD002$flow < decile, "low flow", "")
head(CAN01AD002[which(CAN01AD002$comment == "low flow"),])
```

### Loops

Loops give programming its power by repeating a task. No more manual repetition! (or at least less of it).

There are 3 types of loops in **R**. 

* the **for** loop will repeat a task *for* the items it is given, then stop on its own
* the **while** loop will repeat a task *while* a given condition is true, and will break once that condition is no longer true
* the **repeat** loop will do just that until the loop is broken with a command such as **break**; note that the **break** command can be used to stop any of the loops above

```{r loop types}
# for loops
for (i in 1:3) {
  print(i)
}

# while loop
i = 1
while (i < 4) {
  print(sprintf("%i is still less than 4",i))
  i = i + 1
}

# repeat loop
i = 1
repeat {
  print(i)
  if (i >= 3) {
	break
  }
  i = i + 1
}
```

Loops in R are generally considered to be slow. However, recent tests
have shown that they have improved, and that they are sometimes actually faster than using the `apply` functions, which are
often used instead of loops: 
https://lorentzen.ch/index.php/2022/02/19/avoid-loops-in-r-really/.
That said, it's best to avoid loops if there is a built-in (vectorized) function which will
do the same thing.

This example shows which method is faster to get the square root of a million numbers
by calculating the execution time for each.

```{r loop speed check example, eval=TRUE, include=TRUE}
d1 <- as.numeric(seq(1, 1e6)) # create a vector of 1 to 1 million
d2 <- vector(length = 1e6)    # create empty vector to hold results in loop

# take the sqrt of each element in the vector
Sys.sleep(1) # pause for 1 second to let CPU settle
system.time(for (i in 1:length(d1)) {d2[i] <- sqrt(d1[i])})

Sys.sleep(1) # pause for 1 second to let CPU settle
system.time(d3 <- sqrt(d1))
```


## Simple plotting

What fun is data if you can't visualize it? Let's plot the `yearly_max` data frame
that we created above. Because we are going to use the first column as the x values
and the second column as the y values, we can use the `plot` function defaults.

```{r plot function, echo=TRUE, eval=TRUE}
plot(yearly_max)
```

This plot is ok, but we can do better. 
We can add lines connecting the points and set its colour and width.
```{r update colours in plot, echo=TRUE, eval=TRUE}
plot(yearly_max)
lines(yearly_max, col = 'red', lwd = 2)
```

What about the axis labels? We can override the defaults and add more descriptive label and a title.
```{r add legend to plot, echo=TRUE, eval=TRUE}

plot(yearly_max, xlab = 'Year',ylab = 'Discharge (m³/s)',
     main = 'WSC Gauge 01AD002 annual max. daily discharge')
lines(yearly_max, col = 'red', lwd = 2)

```

Want to output your plot to a file? Try this out.
```{r save a plot, eval=FALSE}
# start a png file of the given name, all subsequent plotting is directed to the file
png("01AD002_annual_max_daily.png") 

plot(yearly_max, xlab='Year',ylab='Discharge (m³/s)',
     main='WSC Gauge 01AD002 annual max. daily discharge')
lines(yearly_max, col='red', lwd=2)

dev.off() # release focus from the png file, return to RStudio

```

You should see an image file (.png) in the current folder (i.e. the one shown by the **getwd()** command).

Note that if you ever need to change the active directory you are in, you can use the **setwd** function. This may be needed if you want to work in a specific directory to conveniently load data during your session. 

\newpage
# More Advanced Applications in R {#advancedR}

Let's take a look at some of the more useful packages and functions in R, as it relates to common tasks and usages of data.

## S3 and S4 objects

**R** was derived from a closed-source program called "S-Plus". One of the features that
**R** inherited is that functions can create their own types of objects, which have
their own ways of being formatted for printing and plotting. Creating these objects (of types S3 and S4)
is a _very_ advanced topic. For now, you just need to know that they exist!

When you do a linear regression of one variable against another (using the `lm` function), the result is an S3 object. So when you `print` and `plot` the regression, you get customized outputs. Note that
because the `plot` function returns 4 different plots, you will be prompted to
hit [Return] to see each one (if you type the command at the console) or
to click on the figures (if you run the chunk).

```{r lm_example1, echo=TRUE, fig.width=4, fig.height = 3}
set.seed(100)       # set random seed so will always get same result (useful for teaching)
a <- c(1, 2, 3, 4, 5)
b <- a + runif(5)   # adds randomness
plot(a, b)
fit <- lm(b~a)
print(fit)
summary(fit)        # shows more information including R²
plot(fit)
```
If you look in the **Environment** tab, you can see that the variable
`fit`, which holds the result of the linear model fit, is a list.

To extract the slope and intercept of the fitted linear model,
you can use the `coefficients` function. To extract R², you need
to use the `summary` function shown above.

```{r lm_example2, echo=TRUE}
# extract slope and intercept
intercept <- coefficients(fit)[1]
slope <- coefficients(fit)[2]
r2 <- summary(fit)$r.squared
# print values
cat("intercept = ", intercept, "slope = ", slope, "R² =", r2)
```

## tidyverse

The **tidyverse** is a collection of **R** packages developed by Hadley Wickam, among
others. They have changed the way **R** is used. Although they take some time to learn,
the packages make **R** much more powerful.


### ggplot2

**ggplot2**, which is part of the **tidyverse**,
gets its name because it incorporates a "Grammar of Graphing". It generates plots in a fundamentally different way than the base plot. With base plots, the plot is a static image that needs to be redrawn if changes are required. In contrast, the `ggplot` function returns an object, which contains all of the data and plot formatting used to generate the plot. The `ggplot` object can then be stored, recalled and plotted, or further modified. Although **ggplot2** works
very differently from the base plot, and takes a while to learn, it is much more powerful, and
allows you to create more attractive plots very easily.

Fortunately, there is a lot of help available to learn **ggplot2**, including a dedicated
web site (https://ggplot2.tidyverse.org/reference/) and even a free book: https://ggplot2-book.org/.

The basic components of a **ggplot2** plot are:

* the `ggplot` call;
* supplying data, usually as a data frame or similar format;
* an aesthetic, i.e. mapping of x, y, and any groupings to the plot;  and
* the format of the plot, i.e. points, lines, etc.


```{r ggplot2 example 1}
library(ggplot2)
p1 <- ggplot(CAN01AD002, aes(date, flow)) +
    geom_line()
p1
```

change axis labels and sizes and colour scheme
```{r ggplot2 example 2}
p2 <- p1 + theme_classic(14)+
     xlab("")+
     ylab("Daily discharge (m³/s)")

p2
```

Examples of more complex plots that can be made using `ggplot2` can be found on the r-statistics.co website: <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>.

You can automatically add regression lines to plots. Saving your graph is a _lot_
easier with **ggplot2** as it has a function `ggsave` which does all the work. The
`width` and `height` are in inches by default, but you can use other units as well.

```{r}
p3 <- ggplot(yearly_max, aes(year, flow)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("") +
  ylab("Maximum daily discharge (m³/s)")
  
p3
graph_file <- "01AD002_annual_max_Q.png"
ggsave(graph_file, width = 8, height = 5)
```

## Functions
If you are doing something repeatedly (like more than 3 times) it's
a good idea to create a function to do it. You can create a function
within your notebook, or you can store it in its own .R file, so that it
can be used in other notebooks.

To create a function, you use the `function` command. You need to specify the
function parameters (which can have default values). Use 
`return` to specify the value that the function will return.

Here's a function that converts from Fahrenheit to Celsius.

```{r}
f2c <- function(temp_f){
  temp_c <- (temp_f - 32) / 1.8
  return(temp_c)
}
```

To use your function, just call it like any other.
```{r}
temps_f <- seq(-40, 100, by = 20)
f2c(temps_f)
```


You can create a file to hold this function by clicking on
**File | New File | R Script**. It creates an empty file.

Now copy the function definition statement above and paste it
into the new R Script.

Save the file clicking  **File | Save**. Give the file a
name ending in ".R". In this case call it "f2c.R".

To load in the function, you can use the command

```{r}
source("f2c.R")
```

You will now see the function `f2c` listed in the Environment tab.


\newpage
# Challenge exercises {#challenges}

## Challenge 1

Using the `ifelse` function, create a new column "hydroyear"
in the CAN01AD002 data frame. If the month is 10 or greater,
the hydro year is one greater than the year. Otherwise, it is the
current year. Hint: you can use the `format` command to get the month
from the date.

## Challenge 2

Get the mean daily discharge for each hydro year. Plot the values.

## Challenge 3

Classify the flows by quarter: q1 = Jan-Mar, q2 = April-June,
q3 = July-September, q4 = October-December.
Plot the flows facetted by each quarter using **ggplot2**
with the function `facet_wrap`.

